---
title: "Point count formatting for BIOL314"
format: pdf
editor: visual
knitr:
  opts_chunk: 
    root.dir: "/Users/aaronskinner/Library/CloudStorage/OneDrive-UBC/Grad_School/PhD/TA_LDP/BIOL314/Biol314_Spring2025_Repo/"
    comment: "#>"
    collapse: true
editor_options: 
  chunk_output_type: console
---

# Goal

Data wrangling to get my bird point count data set into a format suitable for BIOL314, where we will do community biodiversity analyses.

Load libraries

```{r}
#| label: Libraries
#| message: false
library(vegan) 
library(tidyverse)
library(janitor)
library(xlsx)
library(ggpubr)
library(cowplot)
ggplot2::theme_set(cowplot::theme_cowplot())
```

I am modeling my data formatting off of the stock datasets in vegan

```{r}
#| label: Vegan-datasets
data(BCI) 
data(BCI.env) 
dim(BCI) # 50 plots, 225 species 
dim(BCI.env) # 50 sites, 9 covs

data(mite) 
dim(mite) #70 sites, 35 species
```

## Create a reduced set of point counts for Biol 314 class

### Site covariates

I am starting with the site covariates as I'm pulling point counts within a single Ecoregion, by certain data collectors, and that have habitat covariate information available

While not best practice, I'm uploading outputs from Rdata files that were created in another R project.

```{r}
#| label: upload-Rdata
load("Rdata/Biol314_inputs_01.06.25.Rdata")
```

NOTE:: I've included the data collector CIPAV at present, but they followed a different methodology so it could make sense to remove them. I'd be interested in keeping them in though if possible!

```{r}
#| label: subset-ids
ids <- Pc_hab %>% filter(Ecoregion == "Piedemonte" & Uniq_db %in% 
                           c("Ubc mbd", "Cipav mbd", "Gaica mbd", "Gaica distancia")) %>% 
  distinct(Id_muestreo, Habitat_cons) %>% # Habitat_cons = habitat consolidated 
  filter(!is.na(Habitat_cons)) %>% 
  arrange(Id_muestreo) %>%
  pull(Id_muestreo) 
length(ids)
ids[1:6]
```

We have `r length(ids)` point counts that we'll include in the dataset.\
\
Covariates -- 1 row per point count

```{r}
#| label: site-covs
Site_covs314 <- Site_covs_df %>% filter(Id_muestreo %in% ids) %>% 
  arrange(Id_muestreo) 
head(Site_covs314)

#Export
Site_covs314 %>% as.data.frame() %>% 
  write.xlsx("Derived/Site_covs314.xlsx", row.names = F)
```

## Bird point count data

**Methodology:** 10 minute 50m radius point counts\
**Data sets included:** Ubc, CIPAV, Gaica MBD, and Gaica distancia for now

```{r}
#| label: Mean-counts-pivot

# Define function for formatting col names
Cap_snake <- function(df){
  df <- df %>% clean_names(case = "snake")
  names(df) <- str_to_sentence(names(df))
  return(tibble(df))
}

Birds_wide <- Birds_fin %>% filter(Id_muestreo %in% ids) %>% 
  distinct(Uniq_db, Id_muestreo, Nombre_ayerbe, Count) %>% 
  reframe(Count_mn = round(mean(Count, na.rm = TRUE), digits = 0), 
          .by = c(Id_muestreo, Nombre_ayerbe)) %>% 
  pivot_wider(id_cols = Id_muestreo, names_from = Nombre_ayerbe, 
              values_from = Count_mn, values_fill = 0) %>% 
  Cap_snake() # Custom function
dim(Birds_wide)
```

`r nrow(Birds_wide)` sites and `r ncol(Birds_wide)` species. Notice this is 2 fewer rows than in the `Site_covs314` data frame. There were no birds observed at these 2 point counts, but these 0s are valuable data. We need to add them back in and replace the NAs with 0s

```{r}
#| Label: Add-sites
Meta314 <- Birds_wide %>% right_join(Site_covs314[c("Id_muestreo")]) %>%
  mutate(across(everything(), ~ replace_na(., 0)))
dim(Meta314) #Better!
```

## Summarize

Summarize data in various ways

```{r}
#| label: Summary
summary_meta <- Meta314 %>% rowwise() %>% 
  mutate(Original_values = list(c_across(where(is.numeric))), # Cache original numeric values
          Spp_obs = sum(Original_values > 0), 
          Max_count = max(Original_values), 
          N_zeros = sum(Original_values == 0), 
          Tot_ind = sum(Original_values) ) %>% 
  distinct(Id_muestreo, Spp_obs, Max_count, N_zeros, Tot_ind) %>% 
  arrange(desc(Spp_obs)) %>% 
  ungroup() 
```

Distribution of summary values (each row is a point count)

```{r}
#| label: Density-plots
names <- c("Species observed", "Max count", "Number of zeros", "Total individuals")
density_plots <- map2(.x = select(summary_meta, -Id_muestreo), .y = names, \(col, name){ 
  ggplot(data = summary_meta, aes(x = col)) + 
    geom_density() + 
    labs(title = name, x = NULL) }) 
ggpubr::ggarrange(density_plots[[1]], density_plots[[2]], 
                  density_plots[[3]], density_plots[[4]])
```

Export bird Excel file

```{r}
#| Label: Export-excel
Meta314 %>% arrange(Id_muestreo) %>% as.data.frame() %>% 
  write.xlsx("Derived/Meta_314.xlsx", row.names = F)
```
